---
title: "Total Deviation Index (TDI) and Coverage Probability (CP)"
author: "Alison Telford"
date: '2023-01-10'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE, warning = FALSE)
list.of.packages <- c("tidyverse", "devtools", "kableExtra", "readxl")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)){install.packages(new.packages)}
suppressMessages(invisible(lapply(list.of.packages, require, character.only = TRUE)))
```

# Definitions

## Total Deviation Index (TDI)

The Total Deviation Index (TDI) describes a boundary, $\kappa$, such that 
$100p\%$ of observations are less than $\kappa$.

## Coverage Probability (CP)

The Coverage Probability (CP) can be considered the inverse of the TDI. It is 
the proportion, $p$, of observations that are less than a pre-specified 
boundary $\kappa$.

# Using the TDI and CP to Assess Agreement

In order to use the statistics for assessing agreement, acceptance criteria
is required. The acceptance criteria for the TDI is determined a priori based on 
biologically and analytically relevant criteria. 

Parameter $p$ is usually chosen to be 0.90 and can be used as an input in the 
TDI calculation to calculate the boundary $\kappa$ which can then be compared
to the acceptance criteria. Usually, if both $\kappa$ and it's upper 95\% 
confidence limit is less than the acceptance criteria, agreement can be 
concluded.

The acceptance criteria can also be used as an input in the calculation of CP 
to calculate $p$. If the value of $p$ and it's lower 95\% confidence limit is
greater than 0.90, agreement can be concluded.

Note that, one can choose to calculate either the TDI or CP to assess agreement
as if one concludes agreement, the other will too. Although it seems common to
report both statistics.

# R Code

```{r, echo = T}
Agreement = function(df, p, k, alpha=0.05, space = 'arithmetic'){
  # Calculate the differences
  diffs = df[,2] - df[,1]
  if (space == 'Percentage'){
    diffs = 100*(diffs/((df[,1] + df[,2])/2))
  }
  n = length(diffs)
  e2 = sum(diffs^2) / (n - 1)
  tdi = qnorm(1 - ((1 - p) / 2)) * sqrt(e2)
  if (e2 != 0){
    w = log(e2)
  }else{
    return(c(tdi, NA, NA, NA))
  }
  means = df %>% 
    summarise(across(everything(), mean))
  mu_d = means[,1] - means[,2]
  d2 = mu_d^2
  if (n < 4){
    return(c(tdi, NA, NA, NA))
  }
  sd2 = (sum(diffs^2) / n - d2) * (n / (n - 3))
  d2_s2 = (d2 / sd2)
  s_w = sqrt(2 * (1 - ((d2^2) / (e2^2))) / (n - 2))
  w_u = w + qnorm(1 - alpha) * s_w
  e2_u = exp(w_u)
  tdi_u = qnorm(1 - ((1 - p) / 2)) * sqrt(e2_u)
  cp = pchisq((k ** 2) / sd2, 1, d2_s2)
  sd = sqrt(sd2)
  kpm = (k + mu_d) / sd
  kmm = (k - mu_d) / sd
  s_t = (
    (
      ((dnorm(-kpm) - dnorm(kmm))^2) +
        (((kmm * dnorm(kmm)) + (kpm * dnorm(-kpm)))^2) / 2
    ) /
      ((n - 3) * (cp^2) * ((1 - cp)^2))
  )
  s_t = sqrt(s_t)
  t = log(cp / (1 - cp))
  t_l = t - qnorm(1 - alpha) * s_t
  cp_l = exp(t_l) / (1 + exp(t_l))
  return(c(tdi, tdi_u, cp, cp_l))
}
```

# Example

The industry standard for measuring liver inflammation is performed using 
medical device $A$. A new medical device, $B$, has been created which also 
measures liver inflammation, but measures it in a more cost effective way. In
order for the medical device to be accepted into a clinical care pathway, it
needs to measure liver inflammation as well as the industry standard. You 
choose to show that medical device $B$ is equivalent to medical device $A$ at
measuring liver inflammation.

It is decided that equivalence between devices will be concluded if the TDI
is less than 30ms.

A total of 100 individuals have had their liver inflammation measured using
both medical device $A$ and $B$. To test for equivalence between the medical
devices, an acceptance criteria of 30ms on the TDI is used.

```{r, echo = T}
set.seed(1)
# Create some dummy data
A = rnorm(100, 800, 50)
B = A + rnorm(100, 0, 10)
# Create a data frame containing the data
df = data.frame(A, B)
# Calculate the TDI and CP
TDI = Agreement(df, 0.9, 30)[1]
TDI_upper = Agreement(df, 0.9, 30)[2]
CP = Agreement(df, 0.9, 30)[3]
CP_lower = Agreement(df, 0.9, 30)[4]
```

Running the above code gives:

- TDI = `r TDI`
- TDI Upper 95\% confidence limit = `r TDI_upper`
- CP = `r CP`
- CP Lower 95\% confidence limit = `r CP_lower`.

Thus as both the TDI and upper 95\% confidence limit of the TDI is less than 30,
agreement can be concluded. Equally because both the CP and lower 95% confidence
limit of the CP is greater than 0.90, agreement can be concluded.